### 2017-11-10 vipassana-for-hackers.tex
v1.2 draft complete

\usepackage{hyperref}
\usepackage{graphicx}
\usepackage{subcaption}
\usepackage{float}
\usepackage{enumitem}
\setlist{noitemsep}

@@

\begin{titlepage}
   \vspace*{\stretch{1.0}}
   \begin{center}
     \Huge\textbf{Vipassana for Hackers}\\
     \Large\textbf{Instigating the Curious}\\
     \vspace{5cm}
     \large\textit{Steven Deobald}\\
     \large\textit{Version 1.1}\\
     \large\textit\today
   \end{center}
   \vspace*{\stretch{2.0}}
\end{titlepage}

@@

This paper will endeavour to avoid falling into the trap of ``Buddhist Guys'' (a mildly derogatory term coined by a serious meditator friend of mine). You might be familiar with phrases commonly attributed to this archetype such as ``When the student is ready, the teacher will appear.'' Mysticism, mystery, and mystique are to be avoided. If I fall into this trap, email me: steven@nilenso.com

@@

@@

The hacker community has been on the verge of this exploration for quite some time. The easiest and most recent examples of this are no doubt the works of Douglas Hofstadter. ``Gödel, Escher, Bach: An Eternal Golden Braid'' captured the imaginations of the hacker community and grew from esoteric pub conversation piece to dinner party staple, around the same time as the term ``blogging'' evolved from portmanteau to an accessible part-time career for stay-at-home parents. The term ``strange loop'', coined in this book, has since become a mild obsession of philosophical and artistic computer scientists, linguists, and mathematicians, even inspiring the namesake software conference held in St. Louis. Hofstadter's more cohesive ``I Am a Strange Loop'' tackles the problem of consciousness more succinctly. However, other than directing the reader to Zen \footnote{Zen is relevant because the practice of Zazen (Zen meditation) is a nervous-system-based meditation similar to Vipassana. The reason I do not discuss Zazen in detail or compare it to Vipassana in this paper is because Zazen, in my experience, is not systematic. The subject of inspection — the nervous system — remains, but Zen teachers tend to avoid explicit instruction, leaving the student to explore on her own. The outcomes, therefore, appear similar but watered down.} at multiple points, these books do not invite the reader to explore these concepts directly. There is no hacking. A thorough read of GEB or IAaSL is akin to reading all four volumes of Knuth's ``The Art of Computer Programming'' without having ever touched a computer or to reading ``The Joy of Cooking'' cover to cover without having ever set foot inside a kitchen.

@@

Eric Raymond described the exercise of LISP mastery to be ``profound enlightenment'' \cite{hackerhowto} and while we can probably agree to ignore the hyperbole, there is something to be said about a language which begins as a calculus and nearly manages to escape formal grammar entirely. LISP has some very interesting properties: Homoiconicity, metaprogramming, self-mutation and self-replication, recursion and trampolining, and — in the case of modern LISPs and other languages with strong concurrency models — hardened notions of immutability and the reification of time.

Combined, these properties tend to blur the lines of compile time and run time, of code and data, of execution and interpretation. Decades ago, thanks to these ``strange loop'' properties of LISP, it was thought to herald the dawn of the Age of Artificial Intelligence. While that did not happen it left us with a fascinating building material: LISP appears to have no upper-bound. It is ``turtles all the way up.'' LISP is not ``turtles all the way down'', as some would describe it. Even if written on a LISP Machine, there are still paths of wires and electrons which need to be dealt with. These are not written in LISP.

Vipassana would appear to be ``turtles all the way down'' in precisely this sense. Vipassana is the process of digging through your consciousness using consciousness itself. At no point do you reach the machine beneath the machine: ``Aha! I found it! The Consciousness Machine. Now I can do away with the meditation and just read the manufacturer's open specification and I'm sorted.'' The machine at the bottom of your consciousness is a strange loop. Having a look at this strange loop is a lot more fun than reading about it.

@@

On the first day, your attention might look something like Figure~\ref{fig:wandering-attention}, where the wavy pink cloud represents any attention spent on any part of the body other than the nostrils or spent ``outside'' the body on thoughts and emotions. By the fourth day, your attention will often take a shape something like Figure~\ref{fig:narrow-attention}. Over the course of the first three days, the instructions narrow down one's attention from the nose-and-lips triangle from Figure~\ref{fig:burmese-anapana-triangle} to a finger-tip-sized area wherever the breath can be felt beneath the nostrils.

@@

If we revisit the notion that ``meditation is a thing'', we can see a specific example in Anapana with the help of the sense map in Figure~\ref{fig:anapana-sense-map}. Where mindfulness and violin, albeit useful physical and mental practices, are not meditation, we can see from the sense map that Anapana \textit{is} meditation. It is the practice of narrowing one's attention, repeatedly, toward a specific sense (in this case, touch) at a specific time (now) in a specific place (just beneath the nostrils). As our skill in Anapana improves, we come nearer and nearer to the focus-oriented meditation ideal of a single point of awareness with unwavering focus.

@@

In Figure~\ref{fig:anapana-sense-map}, we can see our attention is (realistically) not perfectly tuned and narrowed to a tiny spot under our nose. We are still likely to hear sounds around us on the third day of the course, especially if the sound is monkeys clambering over the roof of the meditation hall.

@@

Through this process, attention will move from the sense of touch to all the internal senses and the sense map will look more like Figure~\ref{fig:vipassana-sense-map}.

\begin{figure}[h]
  \centering
  \includegraphics[width=8cm]{images/ma-vipassana1.png}
  \caption{Vipassana sense map.}
  \label{fig:vipassana-sense-map}
\end{figure}

I should note that I have never successfully penetrated my muscles or inner tissues much. The average new meditator is likely to feel a great many fascinating sensations on her fingertips, palms, arms, toes, and legs. The torso, since it is surrounded in longer nerve fibres, will take longer to observe in the same way. \textit{[citation needed]}

@@

As a Canadian who spent little (if any) of my childhood sitting cross-legged on the floor, Figure~\ref{fig:burmese-vipassana-pain} roughly summarizes most of my early meditation experiences, where the purple dots are intense pain and the pink dot continues to be my awareness. At the peak of my meditation practice toward the end of a course, I would find that if my attention were somewhere away (say, my neck) from the painful area of my body (say, my knees), I may not feel any pain at all. As I moved my attention around, the pain in my knees would come back as I approached them, but every time I could sustain focus on another part of my body, the pain would appear to decrease with every return visit to my knees. If I can remain calm and focused while away from my knees, my knees' pain seems to take care of itself. If I can remain calm and focused while my attention moves across my knees, the situation continues to improve.

@@

These ``thoughts'' are not normal thoughts in any sense of the word you are currently familiar. These ``thoughts'' are an entirely new form of brain activity caused by your constant prodding of the nervous system through Vipassana. Everything in blue, unfortunately, falls into the ``explaining the concept of `white' to a birth-blind man'' category of unexplainable sensory input.

What is meant by ``prodding our nervous system'' and what does this have to do with brain activity? Our second suspension of disbelief was of the deep and intense connection between the brain the nervous system. Alternatively, we can think of the pair as the brain and the body. Alternatively again, we can think of them as a single system.

To think of the brain and body as one system is not incorrect but defeats the model we're drawing out at present. For the brain and nervous system to be ``mutually recursive'', they must be separate processes or entities. What is meant then by stating that the brain and nervous system are mutually recursive?

This statement is meant operationally, not existentially. We do not mean to say that the brain could not live without the nervous system (or vice-versa) but that every event — every operation — occurring in the nervous system has a corresponding event in the brain, and so forth. This mutually recursive process will eventually wear itself out. If it did not, the advent of slight anger would mean we would, from that moment onward, just get angrier and angrier for the rest of our lives. However, during the process of mutual recursion we can be feeling pain in a knee, to which our emotional response is ``argh, I hate pain'', which makes us angry, which clenches up our physical systems, which makes the pain worse, and so on.

Vipassana is the process of unwinding the stack: If I can pay attention to my aching knee, remain calm and continue to observe it, slowly my body unclenches, the pain subsides a little, my emotional response is calmer, and so on.

Blue Bars 2:
The New-New Distractions

Let us quickly revisit our third suspension of disbelief: ``dissolution of the body is a thing.'' Neuroscience tells us some interesting things about the nature of our sense of self and how it relates to our body. From ``Where Am I?'' \cite{whereami}:

``The studies reported that people who had once been able-bodied and then became paralyzed felt… less. Less happy than able-bodied people, less sad than able-bodied people. Just less.''

``Our being is rooted in a body-state. If I were able to remove from your brain, the representation of your body, you would not know that you were you.''

...which would seem in line with experiences described by mastered meditators of having ``no I'' or ``no self'' as their fundamental insightful discovery. It also seems necessary to project our experiences from one 10-day Vipassana course into the future, with respect to the extent to which this practice can take us. If it were stated that we could explore the entirety of the nervous system and its relationship to the brain — but not the gallbladder or spinal cord because those are off-limits — we would have reason to believe that there is a portion of the body and/or nervous system which is exempt from the rules governing these interactions.

@@

This I/O feedback between the brain and body appears to create a completely new kind of distraction. I will attempt to highlight this in the following diagram:
Here, all of the sense doors which we have quiesced return to life, as did ``thought'' before them. However, this new sensory input (which remains a distraction... just a very sophisticated distraction) is unlike most of regular experience. The nearest approximation I can come up with would be the experience of taking salvia, an awful drug which is legally permitted in most countries because it is simply not that enjoyable. While I do not at all recommend smoking salvia, the Wikipedia article [9] is worth a glance. The to relevant items from the list of ``immediate effects'' are:
Past memories, such as revisiting places from childhood memory
Overlapping realities, such as the perception of being in several locations at once

The effects of deep Vipassana meditation are akin to this, in the sense that one may experience, in its entirety, a completely different time or reality. That is where similarities end but the comparison may help those who have experience with hallucinogens.

It is worth reiterating that the purpose of Vipassana is not to experience altered mental states or its potentially hallucinogenic effects. New thought patterns and ``new realities'' are, again, simply distractions from the meditation itself. The meditation remains the systematic observation of the nervous system.Examples

While I would like to avoid descriptions of personal experiences, a few may help highlight the above description of new distracted states of mind.

The non-example:
the resolution to paradox

@@

These would appear to mirror Zen koan practice, where the paradox is presented to the meditator verbally, in advance. No such practice occurs in Vipassana. Paradoxes and their solutions appear to be intrinsic to the human makeup, somehow.

Holding a baby

A friend once had a very salvia-like experience, feeling as though she was teleported to her friend's kitchen, holding her friend's 2-year-old while he cried. The experience was so raw, realistic, and unexpected that she actually opened her eyes and stood up in her meditation cell.

Writing a poem

Often, in early stage meditation, a person will find him or herself distracted by an upcoming task: write this email, call this person, do this chore.

I had a strangely parallel experience toward the end of my second 10-day course. My brain appeared to have written a poem: a visceral, all-encompassing work of art which touched my every sense and filled my mind. The temptation was not to read the poem (since ``I'' had written it) but to write the last line... to take credit for the poem and make it complete. Given that this genuinely felt like the most beautiful thing I had ever created, it was extremely difficult to fight the temptation to finish it and instead continue scanning my lower back with my awareness.

Time slowing down

@@

However, this is not what I am referring to. The experience of time slowing down in very deep meditative states (which I have not experienced a great deal, granted) is that of time slowing down a great deal ...10s or 100s of thousands of years pass before your awareness in a single meditation period.

Time speeding up

The flip side of this, of course, is time speeding up. One sits down to meditate, takes a comfortable position, begins the process in earnest, and BOOM, the gong is going off. The meditation period is over.

Time going backward

Much harder to describe than the experience of time stretching out or time compressing into a single instant is the experience of time moving backward. It would appear, to one's awareness, that cause and effect have been effectively reversed. Each moment takes you a moment (or many) backward in time, and the outcomes of your thoughts or feelings are the source material to their original causes, now future events.

This experience is not necessarily isolated, either. One may experience this as a sort of ``scrobbling'' of time back and forth. This can be quite disorienting and, as distractions go, is extremely difficult to avoid while trying to keep your attention moving over your right nipple.

But that's still the game.

Outcomes

@@

There is a sketch in ``Where Am I?'' \cite{whereami} about an arguing couple who must overcome their emotional temptation to get wound up by anger, ending in this wonderful quote:

``Honey, don’t forget what the half-life is on the autonomic nervous system.''

The autonomic nervous system is what causes the ``winding stack'' of thought and emotion described earlier. The autonomic nervous system is what makes -- and keeps -- you angry. And afraid. And sad. And anxious. And everything else.

@@

I would describe the positive effects of nervous system meditations to live on a spectrum. On one end of the spectrum are the positive, rationality-enhancing qualities of calming one's nervous system with quiet attention. On the other end of the spectrum are the subtler qualities: Meditation seems to make a person more generous, more compassionate, more empathetic. Unlike the trick to healthy rationality — intentional pacifying of the autonomic nervous system — I do not have a pop neuroscience answer as to why this is the case.

I do find the entire endeavour not just fascinating, but endlessly worthwhile. Vipassana is simultaneously the hardest thing I have ever done — and the most valuable.Further Reading

My write-up of my first 10-Day Vipassana course:
http://blog.deobald.ca/2013/02/an-atheists-glimpse-of-god.html

My comparison of meditation techniques and various drugs:
http://blog.deobald.ca/2015/01/drugs-meditation-warnings.html

Appendix A:
Can we ``hack'' Vipassana?

If Vipassana is hacking the mind, is there an opportunity for hacking Vipassana? Can we hack the hacking process? Let's ask Martin Thompson. The following slide is from his 2013 talk ``Performance Testing Java Applications'' \cite{testingjava} and while I have no intention of ostracizing non-computer hackers, software was my career once upon a time and the example is appropriate.

He suggests that any system we intend to observe should be observed through a 3rd party observer by dumping network traffic. For computer systems, this makes a great deal of sense. Were the system under test or the load agents (the Death Star and X-Wings, respectively) to perform the measurements, the measurements and the systems would likely interfere with each other.

Unfortunately, human biology is not so keen on this setup. Even if we had the technology to tap into the interactions of the brain and nervous system (we don't), there is no way to observe the interactions of the brain and nervous system without immediately falling into the Matrix Paradox. How do I know the tool I used is trustworthy? How do I know the observations I made with my eyes are trustworthy? This is all once again external.

It is possible to ``jump out'' of Vipassana momentarily to take stock of the effects of the meditation. Did my hearing just cut out? Did I just force my brain to become consciously aware of my normally-unconscious proprioception pertaining to my right arm? Did I just feel my cardiovascular system in exquisite detail or did I imagine it? …but it is not possible to do so without breaking the meditation for the period of this observation.

@@

>>>>

\usepackage{hyperref}
\usepackage{graphicx}
\usepackage{subcaption}
\usepackage{float}
\usepackage[bottom]{footmisc}
\usepackage{enumitem}
\setlist{noitemsep}

@@

\begin{titlepage}
   \vspace*{\stretch{1.0}}
   \begin{center}
     \Huge\textbf{Vipassana for Hackers}\\
     \Large\textbf{Instigating the Curious}\\
     \vspace{5cm}
     \large\textit{Steven Deobald}\\
     \large\textit{Version 1.2}\\
     \large\textit\today
   \end{center}
   \vspace*{\stretch{2.0}}
\end{titlepage}

@@

This paper will endeavour to avoid falling into the trap of ``Buddhist Guys'' (a mildly derogatory term coined by a serious meditator friend of mine). You might be familiar with phrases commonly attributed to this archetype such as ``When the student is ready, the teacher will appear.'' Mysticism, mystery, and mystique are to be avoided. If I fall into this trap, email me: \href{mailto:steven@nilenso.com}{steven@nilenso.com}

@@

\vspace{1cm}

@@

The hacker community has been on the verge of this exploration for quite some time. The easiest and most recent examples of this are no doubt the works of Douglas Hofstadter. ``Gödel, Escher, Bach: An Eternal Golden Braid'' captured the imaginations of the hacker community and grew from esoteric pub conversation piece to dinner party staple, around the same time as the term ``blogging'' evolved from portmanteau to an accessible part-time career for stay-at-home parents. The term ``strange loop'', coined in this book, has since become a mild obsession of philosophical and artistic computer scientists, linguists, and mathematicians, even inspiring the namesake software conference held in St. Louis. Hofstadter's more cohesive ``I Am a Strange Loop'' tackles the problem of consciousness more succinctly. However, other than directing the reader to Zen\footnote{Zen is relevant because the practice of Zazen (Zen meditation) is a nervous-system-based meditation similar to Vipassana. The reason I do not discuss Zazen in detail or compare it to Vipassana in this paper is because Zazen, in my experience, is not systematic. The subject of inspection — the nervous system — remains, but Zen teachers tend to avoid explicit instruction, leaving the student to explore on her own. The outcomes, therefore, appear similar but watered down.} at multiple points, these books do not invite the reader to explore these concepts directly. There is no hacking. A thorough read of GEB or IAaSL is akin to reading all four volumes of Knuth's ``The Art of Computer Programming'' without having ever touched a computer or to reading ``The Joy of Cooking'' cover to cover without having ever set foot inside a kitchen.

@@

Eric Raymond described the exercise of LISP mastery to be ``profound enlightenment'' \cite{hackerhowto} and while we can probably agree to ignore the hyperbole, there is something to be said about a language which begins as a calculus and nearly manages to escape formal grammar entirely. LISP has some very interesting properties:

\begin{itemize}
\item Homoiconicity: the language is represented in itself and code written in the language is understood by the language
\item Metaprogramming: the language can program itself and understand itself
\item Self-mutation and self-replication: a LISP program can inherently alter its own internal behaviour and easily write entirely new programs
\item Recursion and trampolining: mathematical functional recursion and mutual recursion between two functions are natural operations in LISP
\end{itemize}

Combined, these properties tend to blur the lines of compile time and run time, of code and data, of execution and interpretation. Decades ago, thanks to these ``strange loop'' properties of LISP, it was thought to herald the dawn of the Age of Artificial Intelligence. While that did not happen it left us with a fascinating building material: LISP appears to have no upper-bound. It is ``turtles all the way up.'' LISP is not ``turtles all the way down'', as some would describe it. Even if written on a LISP Machine (a computer which executes LISP directly rather than transforming LISP into machine-level instructions), there are still paths of wires and electrons which need to be dealt with. These are not written in LISP.

Vipassana would appear to be ``turtles all the way down'' in precisely this sense. Vipassana is the process of digging through your consciousness using consciousness itself. At no point do you reach the machine beneath the machine: ``Aha! I found it! The Consciousness Machine. Now I can do away with the meditation and just read the manufacturer's open specification and I'm sorted.'' The machine at the bottom of your consciousness \textbf{is} a strange loop. Having a look at this strange loop is a lot more fun than reading about it.

@@

On the first day, your attention might look something like Figure~\ref{fig:wandering-attention}, where the wavy pink cloud represents any attention spent on any part of the body other than the nostrils or spent ``outside'' the body: on thoughts and emotions. By the fourth day, your attention will often take a shape something like Figure~\ref{fig:narrow-attention}. Over the course of the first three days, the instructions narrow down one's attention from the nose-and-lips triangle from Figure~\ref{fig:burmese-anapana-triangle} to a finger-tip-sized area wherever the breath can be felt beneath the nostrils.

@@

If we revisit the notion that ``meditation is a thing'', we can see a specific example in Anapana with the help of the sense map in Figure~\ref{fig:anapana-sense-map}. Where mindfulness and violin, albeit useful physical and mental practices, are not meditation, we can see from the sense map that Anapana \textit{is} meditation. It is the practice of narrowing one's attention, repeatedly, toward a specific sense (in this case, touch) at a specific time (now) in a specific place (just beneath the nostrils). As our skill in Anapana improves, we come nearer and nearer to the focus-oriented meditation ideal of an unwavering single point of awareness.

@@

In Figure~\ref{fig:anapana-sense-map}, we can see that in reality our attention is not perfectly tuned and narrowed to a tiny spot under our nose. We are still likely to hear sounds around us on the third day of the course, especially if the sound is monkeys clambering over the roof of the meditation hall. By the third day of the course, however, one's attention to sound is likely to be greatly reduced.

For the rest of this paper, make note of the fact that \textit{distraction} is at this stage, and will forever remain, the most important aspect of meditation; it is everything meditation is not, and therefore a significant consideration in all the techniques which follow. Meditation is essentially comprised of only two components: (a) the meditation object and (b) distractions. The activity of meditation is the repeatedly directing of one's attention from the latter to the former.

@@

Through this process, attention will move from the sense of touch to all the internal senses and the sense map will look more like Figure~\ref{fig:vipassana-sense-map-1}.

\begin{figure}[H]
  \centering
  \includegraphics[width=\linewidth]{images/ma-vipassana1.png}
  \caption{Vipassana sense map.}
  \label{fig:vipassana-sense-map-1}
\end{figure}

I should note that I have never successfully penetrated my muscles or inner tissues much. The average new meditator is likely to feel a great many fascinating sensations on her fingertips, palms, arms, toes, and legs. The torso, since it is surrounded in longer nerve fibres, will take longer to observe in the same way. \textit{[citation needed]}\\

@@

As a Canadian who spent little (if any) of my childhood sitting cross-legged on the floor, Figure~\ref{fig:burmese-vipassana-pain} roughly summarizes most of my early meditation experiences, where the purple dots represent intense pain and the pink dot continues to represent my awareness. At the peak of my meditation practice toward the end of a course, I would find that if my attention were somewhere away (say, my neck) from the painful area of my body (say, my knees), I may not feel any pain at all. As I moved my attention around, the pain in my knees would come back as I approached them, but every time I could sustain focus on another part of my body, the pain would appear to decrease with every return visit to my knees. If I can remain calm and focused while away from my knees, my knees' pain seems to take care of itself. If I can remain calm and focused while my attention moves across my knees, the situation continues to improve.

@@

\begin{figure}[h]
  \centering
  \includegraphics[width=\linewidth]{images/ma-vipassana3.png}
  \caption{Attention on a new distraction.}
  \label{fig:vipassana-sense-map-2}
\end{figure}

These distracting new ``thoughts'' are not normal thoughts in any sense of the word you are currently familiar. These ``thoughts'' are an entirely new form of brain activity caused by your constant prodding of the nervous system through Vipassana. Everything in blue, unfortunately, falls into the ``explaining the concept of `white' to a birth-blind man'' category of unexplainable sensory input. Though this brain activity cannot be explained it \textbf{can} be experienced directly and even a first-time meditator is likely to experience such brain activity on her first 10-day course. The blue bar in Figure~\ref{fig:vipassana-sense-map-2} does not represent a goal or form of meditation, nor even a binary state. Rather, if we imagine brain activity on a continuum, worries about taxes might appear on the far left, interesting math and logic problems in the centre, and the ``blue bars'' of unexplainable thought on the right. However fascinating a journey toward this brain activity may be, the thought processes themselves remain nothing more but a distraction from our goal.

What is meant by ``prodding our nervous system'' and what does this have to do with brain activity? Our second suspension of disbelief was of the deep and intense connection between the brain the nervous system. We cannot feel the brain directly --- or at all. Thanks to the deep interconnection between the brain and the nervous system, however, the body acts as a surface to which we have ready access --- and which in turn provides us access to the brain. This brain/body connection can be thought of as \textit{mutual recursion}, as the term is used in the previous section.

That the brain/body connection is \textit{mutually recursive} is meant operationally, not existentially. We do not mean to say that the brain could not live without the nervous system (or vice-versa) but that every event --- every operation --- occurring in the nervous system has a corresponding event in the brain, and so forth. This mutually recursive process will eventually wear itself out. If it did not, the advent of slight anger would mean we would, from that moment onward, just get angrier and angrier for the rest of our lives. However, during the process of mutual recursion we can be feeling pain in a knee, to which our emotional response is ``argh, I hate pain'', which makes us angry, which clenches up our physical systems, which makes the pain worse, and so on.

This stack of mutually-recursive operations is what produces all thought, emotion, and sensation. Vipassana is the process of unwinding the stack: If I can pay attention to my aching knee, remain calm and continue to observe it, slowly my body unclenches, the pain subsides a little, my emotional response is calmer, I am able to remain calm longer, and so on.

\begin{center}
  \Huge{Blue Bars 2:}\\
  \Huge{The New-New Distractions}
\end{center}

Let us quickly revisit our third suspension of disbelief: ``dissolution of the body is a thing.'' Neuroscience tells us some interesting things about the nature of our sense of self and how it relates to our body. From ``Where Am I?'' \cite{whereami}:\\

\textbf{``The studies reported that people who had once been able-bodied and then became paralyzed felt… less. Less happy than able-bodied people, less sad than able-bodied people. Just less.''}\\

\textbf{``Our \textit{being} is rooted in a body-state. If I were able to remove from your brain, the representation of your body, you would not know that you were you.''}\\

...which would seem in line with experiences described by mastered meditators of having ``no I'' or ``no self'' as their fundamental insightful discovery. It also seems necessary to project our experiences from one 10-day Vipassana course into the future, with respect to the extent to which this practice can take us. If it were stated that we could explore the entirety of the nervous system and its relationship to the brain --- but not the gallbladder or spinal cord because those are off-limits --- we would have reason to believe that there is a portion of the body and/or nervous system which is exempt from the rules governing these interactions.

@@

This I/O feedback between the brain and body appears to create \textit{a completely new kind of distraction, which goes well beyond new thought patterns}. An attempt to highlight this form of distraction is given in Figure~\ref{fig:vipassana-sense-map-4}.

\begin{figure}[h]
  \centering
  \includegraphics[width=\linewidth]{images/ma-vipassana4.png}
  \caption{Attention distracted by all-new senses.}
  \label{fig:vipassana-sense-map-4}
\end{figure}

Here, all of the sense doors which we have quiesced return to life, as did ``thought'' before them. However, this new sensory input (which remains a distraction... just a very sophisticated distraction) is unlike most of regular experience. The nearest approximation I can come up with would be the experience of taking salvia, an awful drug which is legally permitted in most countries because it is simply not that enjoyable. While I do not at all recommend smoking salvia, the Wikipedia article \cite{salvia} is worth a glance, since the description of hallucinogens is approachable for anyone. The two relevant items from the list of ``immediate effects'' are:

\begin{itemize}
  \item Past memories, such as revisiting places from childhood memory
  \item Overlapping realities, such as the perception of being in several locations at once
\end{itemize}

The effects of deep Vipassana meditation are akin to this, in the sense that one may experience, in its entirety, a completely different time or reality. A comparable, sober experience might be that of a car crash: participants often feel time slow down or stop, which is sensory input completely detached from the normal flow of consciousness. That is where similarities end but the comparisons may help provide us with tangible, and therefore approachable, examples.

It is worth reiterating that the purpose of Vipassana is not to experience altered mental states. New thought patterns and new realities are, again, simply distractions from the meditation itself. The meditation remains the systematic observation of the nervous system.

\begin{center}
  \Huge{Examples}
\end{center}

While I would like to avoid descriptions of personal experiences, a few may help highlight the above description of new and unfamiliar distracted states of mind.

These examples also serve a secondary purpose. If one searches for ``vipassana'' on YouTube, one will find Goenka's instructional videos, critiques from individuals who could not complete their course, TED talks, and experience reports from individuals who did complete their first course. These experience reports (often titled ``My experience...'') tend to confuse these distracting experiences with the meditation itself. Framing a few examples within this paper should help the reader with this differentiation.

\vspace{1cm}
\begin{center}
  \LARGE{The non-example:}\\
  \LARGE{The Resolution to Paradox}
\end{center}

@@

These would appear to mirror Zen koan practice, where the paradox is presented to the meditator verbally, in advance. No such practice occurs in Vipassana. Paradoxes and their solutions appear to be intrinsic to the human makeup, somehow. While the resolution of paradox may be of acute interest to many hackers, indulging in the consideration of a manifest paradox is as much a tangent and distraction to Vipassana as it is to Zazen.

\vspace{1cm}
\begin{center}
  \LARGE{Holding a Baby}
\end{center}

A friend once had a hallucination-like experience, feeling as though she was teleported to her friend's kitchen, holding her friend's 2-year-old while the child cried. The experience was so raw, realistic, and unexpected that she actually opened her eyes and stood up in her meditation cell.

Such distractions are difficult to overcome, since they appear to replace consciousness and reality itself.\\

\vspace{1cm}
\begin{center}
  \LARGE{Time Slowing Down}
\end{center}

@@

However, this is not what I am referring to. The experience of time slowing down in very deep meditative states is that of time slowing down a \textbf{great deal} ...tens or hundreds of thousands of years pass before your awareness in a single meditation period.

It is difficult to describe how this effect remains nothing but a distraction. However, if meditation remains subdivided into two categories, the meditation object and distractions therefrom, we must accept that even experiences which alter our fundamental perceptions are little more than our brain's creative attempts to distract us from the meditation itself.

\begin{center}
  \Huge{Outcomes}
\end{center}

@@

There is a sketch in \textit{Where Am I?} \cite{whereami} about an arguing couple who must overcome their emotional temptation to get wound up by anger, ending in this wonderful quote:\\

\textbf{``Honey, don’t forget what the half-life is on the autonomic nervous system.''}\\

The autonomic nervous system is what causes the ``winding stack'' of thought and emotion described earlier. The autonomic nervous system is what makes --- and keeps --- you angry. And afraid. And sad. And anxious. And everything else you feel.

@@

I would describe the positive effects of nervous system meditations to live on a spectrum. On one end of the spectrum are the positive, rationality-enhancing qualities of calming one's nervous system with quiet attention. On the other end of the spectrum are the subtler qualities: Meditation seems to make a person more generous, more compassionate, more empathetic. Why would systematic analysis of the autonomic nervous system have these effects? A wide body of research suggests that many forms of meditation, including Vipassana, exhibit these positive effects. Many often identify specifically enhanced activity in the prefrontal cortex \cite{differentialengagement, meditationbrainstructure}, which is identified as the centre of the will to live and altruistic tendencies in the human brain. \cite{personalityneuroscience} However, literature explaining why or how such effects may occur is limited.

Regardless of the mechanism, I do find the entire endeavour not just fascinating, but endlessly worthwhile. Vipassana is simultaneously the hardest thing I have ever done --- and the most valuable.

\begin{center}
  \Huge{Further Reading}
\end{center}

\vspace{1cm}

\begin{itemize}
\item \textit{The Art of Living: Vipassana Meditation As Taught by S. N. Goenka} (Hart, 1987)
\item \textit{An Atheist's Glimpse of God}, my write-up of my first 10-Day Vipassana course:
  \begin{itemize}
  \item \url{http://blog.deobald.ca/2013/02/an-atheists-glimpse-of-god.html}
  \end{itemize}
\item \textit{Drugs, meditation, warnings}, my comparison of meditation techniques and various drugs:
  \begin{itemize}
  \item \url{http://blog.deobald.ca/2015/01/drugs-meditation-warnings.html}
  \end{itemize}
\end{itemize}

\begin{center}
  \Huge{Appendix A:}\\
  \Huge{Can we ``hack'' Vipassana?}
\end{center}

If Vipassana is hacking the mind, is there an opportunity for hacking Vipassana? Can we hack the hacking process? Let's ask Martin Thompson. The following slide in Figure~\ref{fig:load-test} is from his 2013 talk \textit{Performance Testing Java Applications} \cite{testingjava} and while I have no intention of ostracizing non-computer hackers, software was my career once upon a time and the example not only appropriate but reasonably accessible.

\begin{figure}[h]
  \centering
  \includegraphics[width=\linewidth]{images/deathstar1.png}
  \caption{A computer system under load test.}
  \label{fig:load-test}
\end{figure}

He suggests that any system we intend to observe should be observed through a 3rd party observer by dumping network traffic (see Figure~\ref{fig:objective-load-test}). For computer systems, this makes a great deal of sense. Were the system under test or the load agents (the Death Star and X-Wings, respectively) to perform the measurements, the measurements and the systems would likely interfere with each other.

\begin{figure}[h]
  \centering
  \includegraphics[width=\linewidth]{images/deathstar2.png}
  \caption{A load test with an objective, 3rd-party observer.}
  \label{fig:objective-load-test}
\end{figure}

Unfortunately, human biology is not so keen on this setup. Even if we had the technology to tap into the interactions of the brain and nervous system (we don't), there is no way to observe the interactions of the brain and nervous system without immediately falling into the Matrix Paradox (see Figure~\ref{fig:nervous-system-observer}). How do I know the tool I used is trustworthy? How do I know the observations I made with my eyes are trustworthy? This is all once again external.

\begin{figure}[h]
  \centering
  \includegraphics[width=\linewidth]{images/deathstar3.png}
  \caption{The nervous system cannot have an objective observer.}
  \label{fig:nervous-system-observer}
\end{figure}

It is possible to ``jump out'' of Vipassana momentarily to take stock of the effects of the meditation. Did my hearing just cut out? Did I just force my brain to become consciously aware of my normally-unconscious proprioception pertaining to my right arm? Did I just feel my cardiovascular system in exquisite detail or did I imagine it? ...but it is not possible to do so without breaking the meditation for the period of this observation.

@@

\bibitem{salvia}
  \url{https://en.wikipedia.org/wiki/Salvia_divinorum}
  \textit{Salvia divinorum}.
  Wikipedia.

\bibitem{differentialengagement}
  Holzel B.K., Ott U., Hempel H., Hackl A., Wolf K., Stark R., Vaitl D.
  \textit{Differential engagement of anterior cingulate and adjacent medial frontal cortex in adept meditators and non-meditators}
  Neuroscience Letters, 421:16–21, 2007.

\bibitem{meditationbrainstructure}
  Fox, Kieran C.R.; Nijeboer, Savannah; Dixon, Matthew L.; Floman, James L.; Ellamil, Melissa; Rumak, Samuel P.; Sedlmeier, Peter; Christoff, Kalina
  \textit{Is meditation associated with altered brain structure? A systematic review and meta-analysis of morphometric neuroimaging in meditation practitioners}
  Neuroscience \& Biobehavioral Reviews, 43:48–73, June 2014.

\bibitem{personalityneuroscience}
  Colin G. DeYoung, Jacob B. Hirsh, Matthew S. Shane, Xenophon Papademetris, Nallakkandi Rajeevan, Jeremy R. Gray.
  \textit{Testing Predictions From Personality Neuroscience: Brain Structure and the Big Five}
  Psychological Science, 21(6):820–828, June 2010.
